---
title: "Final_Project"
author: "Ananya, Rosie, Sameer"
date: '2025-04-04'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Set working directory
knitr::opts_knit$set(root.dir = "/Users/sameerswarup/Documents/Duke MEM Classes/Spring 2025/Time Series Analysis for Energy/RStudio Outputs/Final_Project")
```

```{r, message= FALSE}
#Load libraries
library(tidyverse)
library(forecast)
#library(tseries)
library(sarima)
library(lubridate)
library(Kendall)
library(outliers)
library(foreign)
library(here)
library(cowplot)
library(kableExtra)
library(smooth)
```

```{r}
#Import datasets
here()
getwd()
alaska_df <- read.csv("../Data/Alaska_combined_output.csv",
                      skip = 2,
                      header = TRUE)
florida_df <- read.csv("../Data/Florida_combined_output.csv",
                      skip = 2,
                      header = TRUE)
maine_df <- read.csv("../Data/Gulf of Maine_combined_output.csv",
                      skip = 2,
                      header = TRUE)
```

```{r}
# Data wrangling needed: create timestamps (sum hourly to daily),
#Sameer comment: Not sure if we can sum, but I can proceed with creating a column that finds the daily average

# Add Wave Energy column: Formula for Wave Energy Flux = Wave Power Density * Wave Energy Period
#Was reading this paper: https://link.springer.com/chapter/10.1007/978-3-319-39889-1_3 and the NOAA explanation on Omni Directional Wave Power. It seems that there is no easy way to compute wave energy flux (so we should not compute it), but that Omni Directional Wave Power is often used as a stand-in as it is correlated with wave energy. The Springer link mentions using wave energy period as another criteria for determining which is a suitable energy site

# Data Exploration: just graph the Wave Energy Flux original series data comparison between the 3 sites.
# Run original TS, ACF, PACF
# Trains models: people tend to use the SARIMA, ARIMA Fourier, STL+ETS, TBATS, 

## state space model can eliminate the impact of outliers
```

```{r}
#Create a column that contains daily mean energy period, wave power and wave height and compute mean wave energy flux
alaska_daily_df <- alaska_df %>% 
              group_by(Year, Month, Day) %>%
              summarise(daily_mean_wave_energy = mean(Energy.Period, na.rm = TRUE),
                        daily_mean_wave_height = mean(Significant.Wave.Height, na.rm = TRUE),
                        daily_mean_wave_power = mean(Omni.Directional.Wave.Power, na.rm = TRUE))

florida_daily_df <- florida_df %>% 
              group_by(Year, Month, Day) %>%
              summarise(daily_mean_wave_energy = mean(Energy.Period, na.rm = TRUE),
                        daily_mean_wave_height = mean(Significant.Wave.Height, na.rm = TRUE),
                        daily_mean_wave_power = mean(Omni.Directional.Wave.Power, na.rm = TRUE))

maine_daily_df <- maine_df %>% 
              group_by(Year, Month, Day) %>%
              summarise(daily_mean_wave_energy = mean(Energy.Period, na.rm = TRUE),
                        daily_mean_wave_height = mean(Significant.Wave.Height, na.rm = TRUE),
                        daily_mean_wave_power = mean(Omni.Directional.Wave.Power, na.rm = TRUE))
```

```{r}
#Create a column that contains monthly mean energy period, wave power and wave height and compute mean wave energy flux
alaska_monthly_df <- alaska_daily_df %>% 
              group_by(Year, Month) %>%
              summarise(monthly_mean_wave_energy = mean(daily_mean_wave_energy, na.rm = TRUE),
                        monthly_mean_wave_height = mean(daily_mean_wave_height, na.rm = TRUE),
                        monthly_mean_wave_power = mean(daily_mean_wave_power, na.rm = TRUE))

florida_monthly_df <- florida_daily_df %>% 
              group_by(Year, Month) %>%
              summarise(monthly_mean_wave_energy = mean(daily_mean_wave_energy, na.rm = TRUE),
                        monthly_mean_wave_height = mean(daily_mean_wave_height, na.rm = TRUE),
                        monthly_mean_wave_power = mean(daily_mean_wave_power, na.rm = TRUE))

maine_monthly_df <- maine_daily_df %>% 
              group_by(Year, Month) %>%
              summarise(monthly_mean_wave_energy = mean(daily_mean_wave_energy, na.rm = TRUE),
                        monthly_mean_wave_height = mean(daily_mean_wave_height, na.rm = TRUE),
                        monthly_mean_wave_power = mean(daily_mean_wave_power, na.rm = TRUE))
```

```{r}
#doing some analysis on the daily and monthly time series to decide what level of granularity might be better for forecasting
#things to look at - volatility (noise), trend/seasonality, ACF patterns
#starting with Maine as our "main" site hehe!

maine_daily_df <- maine_daily_df %>%
  mutate(date = make_date(Year, Month, Day))

maine_monthly_df <- maine_monthly_df %>%
  mutate(date = make_date(Year, Month, 1))

#initial plots
maine_d_plot <- ggplot(maine_daily_df, aes(x = date, y = daily_mean_wave_power)) +
  geom_line(color = "steelblue", alpha = 0.5) +
  labs(title = "Daily Mean Wave Power - Maine", y = "Wave Power (W/m)", x = "") +
  theme_minimal()

maine_m_plot <- ggplot(maine_monthly_df, aes(x = date, y = monthly_mean_wave_power)) +
  geom_line(color = "darkgreen") +
  labs(title = "Monthly Mean Wave Power - Maine", y = "Wave Power (W/m)", x = "") +
  theme_minimal()

plot_grid(maine_d_plot, maine_m_plot, ncol = 1)


#acf plots
maine_d_acf <- ggAcf(maine_daily_df$daily_mean_wave_power) + ggtitle("ACF - Daily Wave Power (Maine)")
maine_m_acf <- ggAcf(maine_monthly_df$monthly_mean_wave_power) + ggtitle("ACF - Monthly Wave Power (Maine)")
plot_grid(maine_d_acf, maine_m_acf, ncol = 1)


```

> Observations: Daily mean wave power appears to be highly volatile and masked by noise, which might make it harder to model directly. On the other hand, monthly mean wave power shows a clear seasonal structure with regular peaks and dips, and consistent annual cycles across decades. 
Moreover, the ACF plot for the daily wave power drops sharply after lag 1, suggesting that past daily values don't carry much signal for future values. For the monthly wave power, the ACF plot shows strong seasonality with the wave pattern. The autocorrelation persists over time, which will be ideal for ARIMA/SARIMA.


```{r}
#Create training and test datasets using the 80-20 rule (for monthly data only)

#Training
alaska_monthly_train_df <- alaska_monthly_df %>%
                          filter(Year < 2004)
florida_monthly_train_df <- florida_monthly_df %>%
                          filter(Year < 2004)
maine_monthly_train_df <- maine_monthly_df %>%
                          filter(Year < 2004)

alaska_monthly_test_df <- alaska_monthly_df %>%
                          filter(Year > 2003)
florida_monthly_test_df <- florida_monthly_df %>%
                          filter(Year > 2003)
maine_monthly_test_df <- maine_monthly_df %>%
                          filter(Year > 2003)
```


```{r}

#converting to time series object
maine_ts_train <- ts(maine_monthly_train_df$monthly_mean_wave_power,
                     start = c(min(maine_monthly_train_df$Year), min(maine_monthly_train_df$Month)),
                     frequency = 12)

maine_ts_test <- ts(maine_monthly_test_df$monthly_mean_wave_power,
                     start = c(min(maine_monthly_test_df$Year), min(maine_monthly_test_df$Month)),
                     frequency = 12)

maine_train_plot <- autoplot(maine_ts_train) +
  ggtitle("Monthly Mean Wave Power - Maine (Training)") +
  xlab("Year") + ylab("Wave Power (W/m)")
plot(maine_train_plot)

#plotting acf and pacf
maine_train_acf <- ggAcf(maine_ts_train) + ggtitle("ACF - Monthly Wave Power (Maine)")
maine_train_pacf <- ggPacf(maine_ts_train) + ggtitle("PACF - Monthly Wave Power (Maine)")
plot_grid(maine_train_acf, maine_train_pacf, ncol = 1)

#decomposing the time series
maine_train_decomp <- decompose(maine_ts_train)
autoplot(maine_train_decomp)

```


```{r}
#start of the forecasting

fore_h = length(maine_ts_test) #forecast horizon

#model 1: seasonal naive on original data (base model)
maine_snaive <- snaive(maine_ts_train, h = fore_h)
autoplot(maine_snaive) + ggtitle("SNAIVE - Maine")


#model 2: SARIMA on original data
maine_sarima_fit <- auto.arima(maine_ts_train)
print(maine_sarima_fit)

maine_sarima_fore <- forecast(object = maine_sarima_fit, h = fore_h)
autoplot(maine_sarima_fore) + ggtitle("SARIMA - Maine")


#model 3: STL decomposition + ETS
maine_stlf_ets <- stlf(maine_ts_train, h = fore_h, method = "ets")
autoplot(maine_stlf_ets) + ggtitle("STL + ETS Forecast - Maine")


#model 4: ARIMA + Fourier terms
k <- 4  #number of Fourier terms kept at 4, as 5 onwards, the RMSE started increasing

maine_arima_fit <- auto.arima(maine_ts_train, 
                             seasonal=FALSE, 
                             lambda=0,
                             xreg=fourier(maine_ts_train, 
                                          K = k))

maine_arima_fore <- forecast(maine_arima_fit,
                           xreg=fourier(maine_ts_train,
                                        K = k,
                                        h = fore_h),
                           h = fore_h)
autoplot(maine_arima_fore) + ggtitle("ARIMA + Fourier Forecast - Maine")

```

```{r}
n_for <- 11 
#play with this, change to 12, 23, 2 and check if the best model is still the same
maine_SSES <- es(maine_ts_train, model="ZZZ", h=n_for, holdout=FALSE)
checkresiduals(maine_SSES)
# forecast and plot
maine_SSES_fore <- forecast(SSES_seas,h=n_for)
plot(maine_SSES_fore)
```
Model own interpretation:
- Residuals kind of fluctuate around 0, but the variance is not very constant, but variance have slightly higher value fluctuations > 0., so some inconsistency.Potential minor Outliers
- ACF plot: no significant spikes outside of the blue lines, which is good
- Histogram of residuals display rough normal distribution around 0, which is good.
- Ljung-Box test: p-value (0.79) much greater than 0.05, so residuals are white noise, good. 


```{r}
SS_seas <- StructTS(maine_ts_train,
                    type="BSM",fixed=c(0.1,0.01,0.3,NA)) 

#this function has convergence issues
checkresiduals(SS_seas)

#Generating forecasts
maine_SS_for <- forecast(SS_seas,h=n_for)
plot(maine_SS_for)
```
This odel without changes from Module 10 was bad, strong autocorrelation and covergence issue, let's change:
```{r}
maine_ts_train_diff <- diff(maine_ts_train, differences = 1)
SS_seas <- StructTS(maine_ts_train_diff, type = "BSM", , fixed = c(NA, NA, NA, NA))  # Let all parameters vary
checkresiduals(SS_seas)

#Generating forecasts
maine_SS_for <- forecast(SS_seas,h=n_for)
plot(maine_SS_for)
```
Revised model: Residual plot improvements (no trends/outliers) suggest:

- The model handles mean and variance reasonably well.

- The seasonal/trend components are likely adequate.

ACF spikes + low p-value imply:

- Short-term dependencies remain unmodeled (e.g., AR/MA effects).

- Seasonal harmonics (higher-frequency cycles) may be missed.

Next steps could be:
- combine the StructTS model with arima layer


```{r}
#summarizing the scores/results from the above models
SNAIVE_scores <- accuracy(maine_snaive$mean,maine_ts_test)  
SARIMA_scores <- accuracy(maine_sarima_fore$mean,maine_ts_test)  
ETS_scores <- accuracy(maine_stlf_ets$mean,maine_ts_test)  
ARIMAF_scores <- accuracy(maine_arima_fore$mean,maine_ts_test)
es_scores <- accuracy(maine_SSES_fore$mean,maine_ts_test)
StructTS_scores <- accuracy(maine_SS_for$mean,maine_ts_test)


scores <- as.data.frame(
  rbind(SNAIVE_scores, SARIMA_scores, ETS_scores, ARIMAF_scores, es_scores, StructTS_scores)
  )
row.names(scores) <- c("SNAIVE", "SARIMA", "STL+ETS", "ARIMA+Fourier", 
                       "ES", "StructTS")

#choose model with lowest RMSE
best_model_index <- which.min(scores[,"RMSE"])
cat("The best model by RMSE is:", row.names(scores[best_model_index,]))

kbl(scores, 
      caption = "Forecast Accuracy for Monthly Wave Power - Maine",
      digits = array(5,ncol(scores))) %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = "hold_position") %>%
  #highlight model with lowest RMSE
  kable_styling(latex_options="striped", stripe_index = which.min(scores[,"RMSE"]))

```


